---
title: "Estimator histories: SCOBI, SCRAPI, EASE, and SCRAPI2"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Rebuild the Lower Granite escapement bootstrap step by step against one
  known-truth simulated fishery, following the estimators from SCOBI and
  SCRAPI to EASE and SCRAPI2.
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# Source the shared helpers from the repository R/ directory. The tutorial runs
# from inst/tutorials/estimator-histories/, so R/ is three levels up; running
# from the repository root also works.
hp <- Filter(file.exists, c("../../../R/tutorial_helpers.R",
                            "R/tutorial_helpers.R",
                            "../../R/tutorial_helpers.R"))
if (length(hp) == 0) stop("tutorial_helpers.R not found")
source(hp[[1]])

# One simulated fishery, preserved across every module.
B_boot    <- 1000L
alpha     <- 0.10
strata    <- 1:3
sim       <- simulate_smolt_season(seed = 6L, B = B_boot)
weeks     <- sim$weeks
rear      <- sim$rear
fish      <- sim$fish
ge_draws  <- sim$ge_draws
gsi_draws <- sim$gsi_draws
truth     <- sim$truth
sim_ad    <- simulate_adult_season(seed = 6L, B = B_boot)
```

## The shared estimator

Both production tools at Lower Granite Dam, EASE for adults and SCRAPI2 for
smolts, run one estimator. Written out, the skeleton is:

1. an observed **count** at the dam;
2. an **inclusion or detection probability** the count is only a fraction of;
3. an **expanded abundance**, count divided by that probability;
4. a **composition** split of the total into groups (rear type, then stock);
5. **uncertainty** propagated to an interval by resampling and posterior draws.

The whole tutorial builds each piece by hand against one simulated smolt season
whose truth we know. Here is the observed season.

```{r show-season, echo=TRUE}
weeks
c(total_smolts = truth$N_total, wild_smolts = truth$WildSmolts)
```

Each week has a raw trap `Tally` (all rear types), a trap sample `rate`, a
guidance efficiency `GE_true`, and a true wild fraction `pWild`. The inclusion
probability for the count is `Ptrue = rate * GE`, and abundance expands as
`count / Ptrue`. The wild total is that abundance times the wild proportion.

### Complete the expansion

Write the plug-in expansion: the total smolts implied by the observed counts at
the point (non-random) inclusion probability.

```{r expand, exercise=TRUE, exercise.lines=5}
# weeks has columns Tally, rate, GE_true. Inclusion prob is rate * GE_true.
# Fill in the expansion and compare it to the truth.
ptrue <- weeks$rate * weeks$GE_true
N_hat <- ______
c(N_hat = round(N_hat), truth = truth$N_total)
```

```{r expand-hint}
# Expanded abundance is the observed count divided by its inclusion
# probability, summed over weeks: sum(count / ptrue).
```

```{r expand-solution}
ptrue <- weeks$rate * weeks$GE_true
N_hat <- sum(weeks$Tally / ptrue)
c(N_hat = round(N_hat), truth = truth$N_total)
```

```{r q-expand}
question(
  "The plug-in estimate for this realized season lands somewhat above the true
   total. What is the single most important reason a ratio estimator like
   count / (rate * GE) is unstable here?",
  answer("The low-GE stratum divides a handful of counts by a probability near
          0.01, so a few counted fish stand in for very many.", correct = TRUE,
         message = "Right: when GE is ~0.08 and the rate is ~0.12, Ptrue is
          ~0.01, and each counted fish expands to ~100. Small count noise
          becomes large abundance noise, which is exactly the right tail the
          later modules propagate."),
  answer("The trap sample rate is biased.",
         message = "The rate enters symmetrically as a known constant here; it
          is the tiny GE, not the rate, that drives the instability."),
  answer("The counts are simulated, so they cannot be trusted.",
         message = "The counts are an honest binomial draw from truth; the
          instability is real, not an artifact of simulation."),
  allow_retry = TRUE
)
```

**Locate this in production.** In `smoltEASE/R/SCRAPI2.R`, the inner estimator
`thetahat()` forms daily abundance as `passage$Tally / passage$Ptrue` and sums
by stratum. On the adult side, `escapeLGD/R/night_fall_reascend_wc_binom.R`
(`expand_wc_binom_night`) expands a window count `wc` by `1 / wc_prop` and by
`1 / (1 - p_night)`. Same skeleton, different names for the probability.

## Build a bootstrap from scratch

A point estimate is one number. The production interval comes from a *compound*
bootstrap: several independent sources of uncertainty stacked into one set of B
replicates. We add them one at a time, then assemble one complete replicate that
carries all of them.

### Stage 1: the plug-in, no uncertainty

```{r stage1, echo=TRUE}
ptrue     <- weeks$rate * weeks$GE_true
est_daily <- round(weeks$Tally / ptrue)   # implied true weekly passage
N_plugin  <- sum(weeks$Tally / ptrue)
round(N_plugin)
```

### Stage 2: the parametric passage-count bootstrap

The count itself is a binomial sample: of `est_daily` true passers, `Tally`
were counted, with probability `ptrue`. Resampling that binomial is the first
uncertainty source.

```{r countboot, exercise=TRUE, exercise.setup = "countboot-setup", exercise.lines=12}
# est_daily and ptrue are defined for you. Draw B binomial resamples of the
# weekly counts and expand each back to a total. Report the bootstrap summary.
B <- 1000
totals <- numeric(B)
for (b in seq_len(B)) {
  cstar <- ______                       # rbinom over the 6 weeks
  totals[b] <- sum(cstar / ptrue)
}
round(boot_summary(totals, truth$N_total), 1)
```

```{r countboot-setup}
ptrue     <- weeks$rate * weeks$GE_true
est_daily <- round(weeks$Tally / ptrue)
```

```{r countboot-hint}
# rbinom is vectorised: rbinom(6, est_daily, ptrue) draws one resampled count
# per week in a single call. Put that inside the loop as cstar.
```

```{r countboot-solution}
B <- 1000
totals <- numeric(B)
for (b in seq_len(B)) {
  cstar <- rbinom(6, est_daily, ptrue)
  totals[b] <- sum(cstar / ptrue)
}
round(boot_summary(totals, truth$N_total), 1)
```

### Stage 3: one coherent GE posterior draw per replicate

GE is not known; it has a posterior. `ge_draws` is a 6-week by B matrix, each
**column** a coherent season-wide draw. The rule that matters: one replicate
uses one whole column, never a different column per week.

```{r geboot, exercise=TRUE, exercise.lines=14}
# ge_draws[, b] is this replicate's GE for all 6 weeks. Build ptrue_b from
# column b, resample the count on that same basis, and expand.
B <- 1000
totals <- numeric(B)
for (b in seq_len(B)) {
  ptrue_b   <- weeks$rate * ge_draws[, b]
  est_b     <- round(weeks$Tally / ptrue_b)
  cstar     <- rbinom(6, est_b, ptrue_b)
  totals[b] <- ______
}
round(boot_summary(totals, truth$N_total), 1)
```

```{r geboot-hint}
# The expansion divides the resampled count by the SAME ptrue_b that generated
# it: sum(cstar / ptrue_b).
```

```{r geboot-solution}
B <- 1000
totals <- numeric(B)
for (b in seq_len(B)) {
  ptrue_b   <- weeks$rate * ge_draws[, b]
  est_b     <- round(weeks$Tally / ptrue_b)
  cstar     <- rbinom(6, est_b, ptrue_b)
  totals[b] <- sum(cstar / ptrue_b)
}
round(boot_summary(totals, truth$N_total), 1)
```

The distribution is now right-skewed: low-GE columns divide by a near-zero
probability and throw a long right tail.

```{r ge-skew, echo=TRUE}
set.seed(1)
totals <- vapply(seq_len(B_boot), function(b) {
  pb <- weeks$rate * ge_draws[, b]
  sum(rbinom(6, round(weeks$Tally / pb), pb) / pb)
}, numeric(1))
c(mean = mean(totals), median = median(totals), skew = skewness(totals))
```

```{r q-ge-skew}
question(
  "Why does adding GE uncertainty widen the interval mostly on the high side?",
  answer("Abundance is proportional to 1/GE, and 1/GE is convex, so low GE
          draws blow up more than high GE draws shrink.", correct = TRUE,
         message = "Yes. This convexity is the same reason E(1/GE) exceeds
          1/E(GE), which the diagnostics module returns to."),
  answer("The count bootstrap is skewed and GE just amplifies it.",
         message = "The binomial count bootstrap is nearly symmetric; the skew
          comes from the 1/GE transform, not the count."),
  allow_retry = TRUE
)
```

**Locate this in production.** `SCRAPI2()` precomputes `ge_day_mat`, a days-by-B
matrix, and inside the bootstrap loop uses `ge_day_mat[, b]` for replicate `b`
(`ptrue_b <- pass[, PASSrate] * ge_day_mat[, b]`). The columns are built by
`generate_ge_draws()` in `smoltEASE`, which selects **whole MCMC rows** so the
GE values on different days stay jointly consistent within a draw.

### Stage 3b: the rear-type stage, and where GE does and does not go

The trap count is all smolts; the wild total is that times the wild proportion.
SCRAPI2 estimates the wild proportion from a rear-type sample (`rear`, each fish
`W` or `H`), weighting by `1/True` where `True = rate * GE`. So GE enters the
wild split too, through `True`. But GE is deliberately **excluded** from the
composition weight `SR = rate * genotype_rate`. The next exercise makes the
distinction concrete: rebuild `True` from a GE column, but leave `SR` alone.

```{r rear, exercise=TRUE, exercise.setup = "rear-setup", exercise.lines=20}
# For replicate b: rebuild True from GE column b, resample rear fish uniformly
# WITHIN each stratum, and estimate the wild proportion by 1/True weighting.
# Then confirm SR (composition weight) is untouched by GE.
b <- 1
True_b <- weeks$rate[rear$week] * ge_draws[cbind(rear$week, b)]   # rate * GE_b
pWild_hat <- numeric(3)
for (h in strata) {
  ri  <- which(rear$stratum == h)
  idx <- ______                                  # uniform resample within stratum
  w_is_wild <- rear$rear[idx] == "W"
  pWild_hat[h] <- sum((1 / True_b[idx]) * w_is_wild) / sum(1 / True_b[idx])
}
c(pWild_hat = round(pWild_hat, 3),
  SR_depends_on_GE = FALSE)   # SR = rate * genotype_rate, GE is not in it
```

```{r rear-setup}
b <- 1
```

```{r rear-hint}
# Uniform within-stratum resample: idx <- sample(ri, length(ri), replace = TRUE).
# ri already holds the row indices for stratum h, so resample those, keeping the
# stratum's sample size.
```

```{r rear-solution}
b <- 1
True_b <- weeks$rate[rear$week] * ge_draws[cbind(rear$week, b)]
pWild_hat <- numeric(3)
for (h in strata) {
  ri  <- which(rear$stratum == h)
  idx <- sample(ri, length(ri), replace = TRUE)
  w_is_wild <- rear$rear[idx] == "W"
  pWild_hat[h] <- sum((1 / True_b[idx]) * w_is_wild) / sum(1 / True_b[idx])
}
c(pWild_hat = round(pWild_hat, 3), SR_depends_on_GE = FALSE)
```

```{r q-rear}
question(
  "Why does SCRAPI2 keep GE inside True (the rear-type weight) but out of SR
   (the composition weight)?",
  answer("Whether wild/hatchery run timing correlates with GE is a real
          empirical question, so True keeps GE; but SR answers 'how many fish
          does this genotyped one represent within what was trapped', which is
          about trap and lab subsampling, not guidance efficiency.",
         correct = TRUE,
         message = "Correct, and this is stated in the SCRAPI2 source: folding
          GE into SR would import the 1/GE tail instability into stock
          composition with no data-driven reason to."),
  answer("It is an arbitrary choice with no consequence.",
         message = "It is deliberate: GE in SR would inflate a few genotyped
          fish on a near-zero-GE day, distorting composition."),
  answer("GE belongs in neither; both should exclude it.",
         message = "True keeps GE on purpose; only SR excludes it."),
  allow_retry = TRUE
)
```

### Stage 4: the full compound SCRAPI2 replicate

Now assemble one complete replicate that carries **every** source at once, and
store all its outputs in the same row: total passage, WildSmolts, stock A, and
stock B. This is the heart of SCRAPI2. Fill in the marked lines.

```{r compound, exercise=TRUE, exercise.setup = "compound-setup", exercise.lines=44}
# Each row is one full replicate. Fill in: the GE column, Ptrue_b, est_daily_b,
# the binomial count resample, the uniform within-stratum resamples, and the two
# inverse-weight estimators (1/True for wild split, 1/SR for stock).
B <- 800
out <- matrix(0, nrow = B, ncol = 4,
              dimnames = list(NULL, c("total", "wild", "stockA", "stockB")))
add_err <- numeric(B)
for (b in seq_len(B)) {
  ge_b      <- ______                              # one whole GE column
  ptrue_b   <- weeks$rate * ge_b
  est_b     <- round(weeks$Tally / ptrue_b)
  cstar     <- rbinom(6, est_b, ptrue_b)           # binomial passage resample
  pass_str  <- tapply(cstar / ptrue_b, weeks$stratum, sum)[as.character(strata)]

  True_b    <- weeks$rate[rear$week] * ge_b[rear$week]   # rate * GE_b per rear fish
  wild_str  <- numeric(3)
  for (h in strata) {
    ri  <- which(rear$stratum == h)
    idx <- sample(ri, length(ri), replace = TRUE)  # uniform within stratum
    pW  <- sum((1 / True_b[idx]) * (rear$rear[idx] == "W")) / sum(1 / True_b[idx])
    wild_str[h] <- pW * pass_str[h]
  }
  total_wild <- sum(wild_str)

  stockA <- 0
  for (h in strata) {
    fi   <- which(fish$stratum == h)
    idx  <- sample(fi, length(fi), replace = TRUE) # uniform within stratum
    st   <- gsi_draws[idx, b]                      # one whole GSI column
    sr   <- fish$SR[idx]
    pA_h <- ______                                 # 1/SR-weighted stock-A proportion
    stockA <- stockA + pA_h * wild_str[h]
  }
  stockB <- total_wild - stockA

  out[b, ] <- c(sum(cstar / ptrue_b), total_wild, stockA, stockB)
  add_err[b] <- (stockA + stockB) - total_wild     # additivity within the row
}
rbind(
  WildSmolts = round(boot_summary(out[, "wild"],   truth$WildSmolts), 1),
  stockA     = round(boot_summary(out[, "stockA"], truth$stockA), 1),
  stockB     = round(boot_summary(out[, "stockB"], truth$stockB), 1))
```

```{r compound-setup}
# global objects (weeks, rear, fish, ge_draws, gsi_draws, truth, strata) are available
```

```{r compound-hint}
# ge_b <- ge_draws[, b] selects one coherent column.
# pA_h <- sum((1 / sr) * (st == "A")) / sum(1 / sr) is the 1/SR-weighted
# stock-A proportion (Horvitz-Thompson), identical in form to the wild split.
```

```{r compound-solution}
B <- 800
out <- matrix(0, nrow = B, ncol = 4,
              dimnames = list(NULL, c("total", "wild", "stockA", "stockB")))
add_err <- numeric(B)
for (b in seq_len(B)) {
  ge_b      <- ge_draws[, b]
  ptrue_b   <- weeks$rate * ge_b
  est_b     <- round(weeks$Tally / ptrue_b)
  cstar     <- rbinom(6, est_b, ptrue_b)
  pass_str  <- tapply(cstar / ptrue_b, weeks$stratum, sum)[as.character(strata)]

  True_b    <- weeks$rate[rear$week] * ge_b[rear$week]
  wild_str  <- numeric(3)
  for (h in strata) {
    ri  <- which(rear$stratum == h)
    idx <- sample(ri, length(ri), replace = TRUE)
    pW  <- sum((1 / True_b[idx]) * (rear$rear[idx] == "W")) / sum(1 / True_b[idx])
    wild_str[h] <- pW * pass_str[h]
  }
  total_wild <- sum(wild_str)

  stockA <- 0
  for (h in strata) {
    fi   <- which(fish$stratum == h)
    idx  <- sample(fi, length(fi), replace = TRUE)
    st   <- gsi_draws[idx, b]
    sr   <- fish$SR[idx]
    pA_h <- sum((1 / sr) * (st == "A")) / sum(1 / sr)
    stockA <- stockA + pA_h * wild_str[h]
  }
  stockB <- total_wild - stockA

  out[b, ] <- c(sum(cstar / ptrue_b), total_wild, stockA, stockB)
  add_err[b] <- (stockA + stockB) - total_wild
}
c(max_additivity_error = max(abs(add_err)))   # ~0: stockA + stockB == WildSmolts every row
rbind(
  WildSmolts = round(boot_summary(out[, "wild"],   truth$WildSmolts), 1),
  stockA     = round(boot_summary(out[, "stockA"], truth$stockA), 1),
  stockB     = round(boot_summary(out[, "stockB"], truth$stockB), 1))
```

```{r q-additivity}
question(
  "The maximum additivity error across all B rows is ~1e-12 (floating point),
   not a statistical near-zero. Why is it exact rather than approximate?",
  answer("Within every replicate stockB is defined as total_wild - stockA, so
          stockA + stockB equals total_wild identically, row by row.",
         correct = TRUE,
         message = "Exactly. SCRAPI2 gets the same guarantee because each row's
          composition proportions sum to 1 by prop.table, so the mean of the
          rows keeps the reported components additive. This is why SCRAPI2
          enforces pointEst = 'mean'."),
  answer("Because the bootstrap has converged with enough replicates.",
         message = "Convergence controls Monte Carlo error, not this identity;
          additivity holds in every single row regardless of B."),
  allow_retry = TRUE
)
```

### Watch the smolt estimator evolve

One row per stage, same simulated season, so you can see each source move the
estimate, the width, and the skew. Coverage is measured across repeated seasons
for the final wild total only.

```{r smolt-history, echo=TRUE}
set.seed(4)
Bh <- 500
count_only <- vapply(seq_len(Bh), function(b) {
  sum(rbinom(6, weeks$Tally, weeks$Ptrue) / weeks$Ptrue)   # GE fixed at point
}, numeric(1))
plus_ge <- vapply(seq_len(Bh), function(b) {
  pb <- weeks$rate * ge_draws[, b]; sum(rbinom(6, round(weeks$Tally / pb), pb) / pb)
}, numeric(1))
# full compound wild total, reusing the compound logic
compound_wild <- function(d, B) {
  vapply(seq_len(B), function(b) {
    pb <- d$weeks$rate * d$ge_draws[, b]
    cs <- rbinom(6, round(d$weeks$Tally / pb), pb)
    ps <- tapply(cs / pb, d$weeks$stratum, sum)[as.character(1:3)]
    Tb <- d$weeks$rate[d$rear$week] * d$ge_draws[cbind(d$rear$week, b)]
    sum(vapply(1:3, function(h) {
      ri <- which(d$rear$stratum == h); idx <- sample(ri, length(ri), replace = TRUE)
      (sum((1 / Tb[idx]) * (d$rear$rear[idx] == "W")) / sum(1 / Tb[idx])) * ps[h]
    }, numeric(1)))
  }, numeric(1))
}
wild_boot <- compound_wild(sim, Bh)
# coverage of the wild total across repeated seasons
cover_wild <- mean(vapply(1:40, function(s) {
  d <- simulate_smolt_season(seed = 100 + s, B = 200L)
  wb <- compound_wild(d, 200L); ci <- quantile(wb, c(.05, .95))
  ci[1] <= d$truth$WildSmolts && d$truth$WildSmolts <= ci[2]
}, logical(1)))
rbind(
  stage_row("count only",        "count",              count_only, truth$N_total),
  stage_row("+ GE posterior",    "count+GE",           plus_ge,    truth$N_total),
  stage_row("+ rear split (wild)","count+GE+rear",     wild_boot,  truth$WildSmolts,
            coverage = round(cover_wild, 2)))
```

## SCRAPI to SCRAPI2

The original SCRAPI and the current SCRAPI2 share this skeleton but differ in
three decisions: whether GE and GSI are fixed or uncertain, how fish are
resampled **within each stratum**, and how the point estimate is defined. The
most consequential is the resampling choice, because it interacts with the
`1/SR` weight already in the estimator.

### Uniform vs SR-weighted resampling, within stratum

The estimator weights each fish by `1/SR`. The question is whether the
*resample* should also use `SR`. Both production tools resample **within each
collapsed stratum**; the only difference is the `prob` argument. Try both and
measure coverage of the true stock-A proportion across many simulated seasons.

```{r resample, exercise=TRUE, exercise.setup = "resample-setup", exercise.lines=24}
# coverage() re-simulates a season each iteration and builds a composition
# interval by resampling WITHIN each stratum, either uniformly or weighted by SR.
coverage <- function(weighted, nsim = 120, B = 250) {
  hit <- logical(nsim)
  for (s in seq_len(nsim)) {
    d  <- simulate_smolt_season(seed = s, B = 1L)
    ff <- d$fish
    th <- numeric(B)
    for (b in seq_len(B)) {
      num <- den <- 0
      for (h in 1:3) {                             # resample within stratum
        fi  <- which(ff$stratum == h)
        idx <- if (weighted)
          sample(fi, length(fi), replace = TRUE, prob = ff$SR[fi])
        else
          sample(fi, length(fi), replace = TRUE)
        num <- num + sum((1 / ff$SR[idx]) * (ff$stock[idx] == "A"))
        den <- den + sum(1 / ff$SR[idx])
      }
      th[b] <- num / den
    }
    ci <- quantile(th, c(.05, .95))
    hit[s] <- ci[1] <= d$truth$pA && d$truth$pA <= ci[2]
  }
  mean(hit)
}
c(uniform = coverage(FALSE), weighted = coverage(TRUE))
```

```{r resample-setup}
# simulate_smolt_season and stats are available
```

```{r resample-hint}
# You do not need to edit anything to see the effect: run it as is. The uniform
# arm keeps 1/SR honest; the weighted arm resamples proportional to SR AND
# divides by SR, so within each stratum the two cancel.
```

```{r resample-solution}
coverage <- function(weighted, nsim = 120, B = 250) {
  hit <- logical(nsim)
  for (s in seq_len(nsim)) {
    d  <- simulate_smolt_season(seed = s, B = 1L)
    ff <- d$fish
    th <- numeric(B)
    for (b in seq_len(B)) {
      num <- den <- 0
      for (h in 1:3) {
        fi  <- which(ff$stratum == h)
        idx <- if (weighted)
          sample(fi, length(fi), replace = TRUE, prob = ff$SR[fi])
        else
          sample(fi, length(fi), replace = TRUE)
        num <- num + sum((1 / ff$SR[idx]) * (ff$stock[idx] == "A"))
        den <- den + sum(1 / ff$SR[idx])
      }
      th[b] <- num / den
    }
    ci <- quantile(th, c(.05, .95))
    hit[s] <- ci[1] <= d$truth$pA && d$truth$pA <= ci[2]
  }
  mean(hit)
}
c(uniform = coverage(FALSE), weighted = coverage(TRUE))
```

```{r q-resample}
question(
  "The uniform arm covers near 90%; the SR-weighted arm covers almost never.
   What has the SR-weighted resample actually estimated?",
  answer("The unweighted sample composition, because resampling proportional to
          SR cancels the 1/SR weight in the estimator, discarding the
          Horvitz-Thompson correction the weight exists to apply.",
         correct = TRUE,
         message = "Correct, and it is why stock A (sampled at a lower rate) is
          badly underestimated. The interval is centered on the wrong quantity,
          so it misses the truth almost every time."),
  answer("The same weighted composition, just with a wider interval.",
         message = "It is not merely wider; it is centered on a different
          quantity (the unweighted proportion), so its coverage collapses."),
  allow_retry = TRUE
)
```

**Locate this in production.** `smoltEASE/R/SCRAPI.R` (original) resamples fish
within week with `sample.int(nwk, replace = TRUE, prob = unlist(justwk$SR))`
while its `thetahat` still divides by `SR` -- the cancellation you just saw.
`smoltEASE/R/SCRAPI2.R` resamples within stratum with
`sample.int(nrow(jw), replace = TRUE)` (no `prob`) and keeps `1/SR` and `1/True`
only inside the estimator. The SCRAPI2 documentation states this directly:
"weighting the resample by them while also applying their inverse inside the
estimator would tend to cancel the intended inverse-probability correction."

### The point estimate is the bootstrap mean over B rows

SCRAPI2 does not report the raw data evaluated once, and it no longer averages a
separate `n_point` set of draws (`n_point` is deprecated). Every one of the B
rows is a full replicate, and the reported `Estimate` is `colMeans(theta.b)`,
the mean of those rows, while `LCI`/`UCI` are percentiles of the same rows.

```{r mean-vs-raw, echo=TRUE}
set.seed(2)
totals <- vapply(seq_len(B_boot), function(b) {
  pb <- weeks$rate * ge_draws[, b]
  sum(rbinom(6, round(weeks$Tally / pb), pb) / pb)
}, numeric(1))
raw_point  <- sum(weeks$Tally / (weeks$rate * weeks$GE_true))
boot_mean  <- mean(totals)                     # colMeans(theta.b) in SCRAPI2
pctl       <- quantile(totals, c(alpha/2, 1 - alpha/2), names = FALSE)
c(raw_point = round(raw_point), boot_mean = round(boot_mean),
  lci = round(pctl[1]), uci = round(pctl[2]))
```

```{r q-meanraw}
question(
  "For a strongly right-skewed cell, the bootstrap MEAN can fall OUTSIDE its own
   percentile interval. Is that a coding bug?",
  answer("No: mean and quantile are different summaries of the same skewed
          distribution, so the mean can sit above the 95th percentile.",
         correct = TRUE,
         message = "Right. SCRAPI2 flags such rows with a message rather than
          hiding them; it signals a heavy right tail to inspect, not an error."),
  answer("Yes: the point estimate must always be inside the interval.",
         message = "Only for symmetric summaries. A mean and a percentile of a
          skewed distribution are genuinely different functionals."),
  allow_retry = TRUE
)
```

## SCOBI to EASE

The adult estimator has its own history. Old SCOBI assumed the counting window
was open the whole time (`wc_prop = 1`); EASE corrects for the window being
closed part of each hour (`wc_prop = 5/6`), expands for nighttime passage, splits
by composition and stock, and adjusts for fallback. Fallback separates two
truths: **ascensions** (fish that climbed the ladder) and **escapement**
(ascensions that did not fall back without reascending).

```{r adult-season, echo=TRUE}
sim_ad$wc                                         # weekly window counts
c(true_ascensions = sim_ad$truth$true_ascensions,
  true_escapement = sim_ad$truth$true_escapement)
```

### SCOBI vs EASE window-count assumption

```{r wcprop, exercise=TRUE, exercise.lines=8}
# The window is open a fraction wc_prop of the time, so the count must be
# expanded by 1/wc_prop. Compute the expanded daytime total under old SCOBI
# (wc_prop = 1) and under EASE (wc_prop = 5/6).
wc_total <- sum(sim_ad$wc$wc)
scobi <- wc_total / 1
ease  <- ______
c(raw = wc_total, scobi = round(scobi), ease = round(ease))
```

```{r wcprop-hint}
# EASE divides by 5/6, which multiplies the count by 6/5: ease <- wc_total / (5/6).
```

```{r wcprop-solution}
wc_total <- sum(sim_ad$wc$wc)
scobi <- wc_total / 1
ease  <- wc_total / (5/6)
c(raw = wc_total, scobi = round(scobi), ease = round(ease))
```

```{r q-wcprop}
question(
  "What does the old SCOBI wc_prop = 1 assumption imply, and why does EASE
   change it?",
  answer("SCOBI assumes every daytime fish is counted (window always open);
          EASE recognizes the window is closed part of each hour, so the raw
          count is only 5/6 of daytime passage and must be expanded.",
         correct = TRUE,
         message = "Correct. wc_prop = 1 leaves out no daytime fish by
          assumption; 5/6 restores the ones passing while the window was shut."),
  answer("SCOBI expands more than EASE.",
         message = "The opposite: wc_prop = 1 applies no expansion at all, so
          SCOBI's daytime total is lower than EASE's."),
  allow_retry = TRUE
)
```

### One row through the whole adult calculation

Now flow one bootstrap row through every adult stage and store the group-specific
escapement. Fill in the marked lines.

```{r adult-full, exercise=TRUE, exercise.setup = "adult-full-setup", exercise.lines=34}
# Each row: window-count binomial bootstrap expanded by 1/wc_prop, nighttime
# expansion, fallback adjustment, then composition + GSI split into stocks.
B  <- 800
wc_prop     <- sim_ad$wc_prop
wc_by_strat <- tapply(sim_ad$wc$wc, sim_ad$wc$stratum, sum)
nd <- sim_ad$night; fd <- sim_ad$fallback; comp <- sim_ad$comp; gsi <- sim_ad$gsi_draws
asc <- esc <- escA <- numeric(B)
for (b in seq_len(B)) {
  wc_b <- vapply(1:3, function(s)
    rbinom(1, round(wc_by_strat[s] / wc_prop), wc_prop) / wc_prop, numeric(1))
  pn_b <- rbinom(3, nd$totalPass, nd$nightPass / nd$totalPass) / nd$totalPass
  ascend_str <- wc_b / (1 - pn_b)                 # window + night = ascensions
  asc[b] <- sum(ascend_str)

  pf_b <- rbinom(3, fd$totalPass, fd$numReascend / fd$totalPass) / fd$totalPass
  esc_str <- ascend_str * (1 - pf_b)              # ascensions - net fallback
  esc[b] <- sum(esc_str)

  a_str <- 0
  for (h in 1:3) {
    ci  <- which(comp$stratum == h)
    idx <- ______                                 # uniform resample within stratum
    st  <- gsi[idx, b]                            # one whole GSI column
    a_str <- a_str + mean(st == "A") * esc_str[h]
  }
  escA[b] <- a_str
}
rbind(
  ascensions = round(boot_summary(asc,  sim_ad$truth$true_ascensions), 1),
  escapement = round(boot_summary(esc,  sim_ad$truth$true_escapement), 1),
  stockA_esc = round(boot_summary(escA, sim_ad$truth$escA), 1))
```

```{r adult-full-setup}
# sim_ad is available
```

```{r adult-full-hint}
# Uniform within-stratum resample of the composition sample:
# idx <- sample(ci, length(ci), replace = TRUE).
# Compare ascensions to true_ascensions (pre-fallback) and escapement to
# true_escapement (after the (1 - pf_b) adjustment).
```

```{r adult-full-solution}
B  <- 800
wc_prop     <- sim_ad$wc_prop
wc_by_strat <- tapply(sim_ad$wc$wc, sim_ad$wc$stratum, sum)
nd <- sim_ad$night; fd <- sim_ad$fallback; comp <- sim_ad$comp; gsi <- sim_ad$gsi_draws
asc <- esc <- escA <- numeric(B)
for (b in seq_len(B)) {
  wc_b <- vapply(1:3, function(s)
    rbinom(1, round(wc_by_strat[s] / wc_prop), wc_prop) / wc_prop, numeric(1))
  pn_b <- rbinom(3, nd$totalPass, nd$nightPass / nd$totalPass) / nd$totalPass
  ascend_str <- wc_b / (1 - pn_b)
  asc[b] <- sum(ascend_str)

  pf_b <- rbinom(3, fd$totalPass, fd$numReascend / fd$totalPass) / fd$totalPass
  esc_str <- ascend_str * (1 - pf_b)
  esc[b] <- sum(esc_str)

  a_str <- 0
  for (h in 1:3) {
    ci  <- which(comp$stratum == h)
    idx <- sample(ci, length(ci), replace = TRUE)
    st  <- gsi[idx, b]
    a_str <- a_str + mean(st == "A") * esc_str[h]
  }
  escA[b] <- a_str
}
rbind(
  ascensions = round(boot_summary(asc,  sim_ad$truth$true_ascensions), 1),
  escapement = round(boot_summary(esc,  sim_ad$truth$true_escapement), 1),
  stockA_esc = round(boot_summary(escA, sim_ad$truth$escA), 1))
```

> **Teaching approximation.** This models fallback as one net rate `p_fall` per
> stratum and multiplies escapement by `(1 - p_fall)`. Production `escapeLGD`
> instead infers fallback and reascension **jointly** with a two-parameter
> binomial likelihood, maximized per bootstrap replicate:
> `optim(par = c(.1, .9), fn = optimllh, ...)` in
> `fallback_reascend_likelihood.R`, driven by `nightFall()`. When spillway data
> are absent it assumes `P(reascend | fallback) = 1`. Composition here is a plain
> stock proportion; production uses the PBT multinomial MLE in
> `composition_estimation_utils.R`. The percentile interval is assembled in
> `apply_fallback_rates()`.

### Watch the adult estimator evolve

```{r adult-history, echo=TRUE}
set.seed(5)
Bh <- 500
wc_by_strat <- tapply(sim_ad$wc$wc, sim_ad$wc$stratum, sum)
nd <- sim_ad$night; fd <- sim_ad$fallback
adult_stage <- function(mode, B) {
  vapply(seq_len(B), function(b) {
    wp <- if (mode == "scobi") 1 else sim_ad$wc_prop
    wc_b <- vapply(1:3, function(s)
      rbinom(1, round(wc_by_strat[s] / wp), wp) / wp, numeric(1))
    if (mode %in% c("scobi", "ease")) return(sum(wc_b))
    pn <- rbinom(3, nd$totalPass, nd$nightPass / nd$totalPass) / nd$totalPass
    ascend <- wc_b / (1 - pn)
    if (mode == "night") return(sum(ascend))
    pf <- rbinom(3, fd$totalPass, fd$numReascend / fd$totalPass) / fd$totalPass
    sum(ascend * (1 - pf))
  }, numeric(1))
}
rbind(
  stage_row("SCOBI window (wc_prop=1)", "window",                adult_stage("scobi", Bh), sim_ad$truth$true_ascensions),
  stage_row("EASE window (wc_prop=5/6)","window",                adult_stage("ease",  Bh), sim_ad$truth$true_ascensions),
  stage_row("+ nighttime passage",      "window+night",          adult_stage("night", Bh), sim_ad$truth$true_ascensions),
  stage_row("+ fallback (escapement)",  "window+night+fallback", adult_stage("fall",  Bh), sim_ad$truth$true_escapement))
```

```{r q-adult}
question(
  "Nighttime passage on the adult side is a bootstrapped binomial rate, not a
   fitted posterior like smolt GE. Which statement is accurate?",
  answer("It is not subject to GE-style model-form or prior bias, but it can
          still be biased if PIT-tagged fish are unrepresentative, detections
          are misclassified, or a stratum pools different passage behavior; the
          binomial bootstrap only captures sampling noise in the tags.",
         correct = TRUE,
         message = "Correct. 'No model to miscenter it' is not 'no bias
          possible' -- representativeness, misclassification, and stratum
          pooling are real threats the binomial interval does not reflect."),
  answer("It carries only sampling noise and can never be biased.",
         message = "Too strong: unrepresentative tags, misclassified
          detections, or heterogeneous strata can bias it even though no fitted
          model can."),
  allow_retry = TRUE
)
```

## Model diagnostics and failure modes

The compound bootstrap has characteristic ways of going wrong. Each below is a
thing to check before quoting an interval.

### E(1/GE) is not 1/E(GE)

```{r jensen, echo=TRUE}
ge_col_means <- rowMeans(ge_draws)                 # per-week posterior mean GE
one_over_mean <- sum(weeks$Tally / (weeks$rate * ge_col_means))
mean_of_one_over <- mean(vapply(seq_len(B_boot), function(b)
  sum(weeks$Tally / (weeks$rate * ge_draws[, b])), numeric(1)))
c(expand_at_mean_GE = round(one_over_mean),
  mean_of_expansions = round(mean_of_one_over))
```

```{r q-jensen}
question(
  "The mean of the expansions exceeds the expansion at the mean GE. Which is the
   honest point estimate for total abundance?",
  answer("The mean of the expansions: abundance is a function of GE, so its
          expectation is the average of the function over the GE posterior, not
          the function evaluated at the average GE.", correct = TRUE,
         message = "Correct. SCRAPI2 computes exactly this by averaging the B
          replicate expansions, which is why it does not plug the posterior mean
          GE into one expansion."),
  answer("The expansion at the mean GE, because it uses one clean GE value.",
         message = "That is the biased one: by Jensen's inequality it
          understates the mean of a convex function of GE."),
  allow_retry = TRUE
)
```

### Mixing GE columns across weeks

The rule from module 2 was: one replicate, one whole GE column. Break it on
purpose and diagnose the damage.

```{r mix, exercise=TRUE, exercise.lines=16}
# WRONG on purpose: draw an INDEPENDENT GE column for each of the 6 weeks within
# a replicate, instead of one shared column. Compare the interval width to the
# coherent version.
B <- 1000
coherent <- mixed <- numeric(B)
for (b in seq_len(B)) {
  pc <- weeks$rate * ge_draws[, b]                       # coherent: one column
  coherent[b] <- sum(weeks$Tally / pc)
  cols <- ______                                         # 6 independent columns
  pm <- weeks$rate * ge_draws[cbind(1:6, cols)]
  mixed[b] <- sum(weeks$Tally / pm)
}
c(coherent_width = diff(quantile(coherent, c(.05,.95))),
  mixed_width    = diff(quantile(mixed,    c(.05,.95))))
```

```{r mix-hint}
# Pick 6 column indices independently: cols <- sample.int(ncol(ge_draws), 6,
# replace = TRUE). ge_draws[cbind(1:6, cols)] then takes week i from column
# cols[i].
```

```{r mix-solution}
B <- 1000
coherent <- mixed <- numeric(B)
for (b in seq_len(B)) {
  pc <- weeks$rate * ge_draws[, b]
  coherent[b] <- sum(weeks$Tally / pc)
  cols <- sample.int(ncol(ge_draws), 6, replace = TRUE)
  pm <- weeks$rate * ge_draws[cbind(1:6, cols)]
  mixed[b] <- sum(weeks$Tally / pm)
}
c(coherent_width = diff(quantile(coherent, c(.05,.95))),
  mixed_width    = diff(quantile(mixed,    c(.05,.95))))
```

```{r q-mix}
question(
  "Mixing independent GE columns across weeks makes the interval narrower. Why
   is that wrong?",
  answer("The weeks share strata and model parameters, so their GE values are
          correlated in the posterior; averaging independent draws cancels that
          shared uncertainty and understates the true spread.", correct = TRUE,
         message = "Correct. This is why generate_ge_draws() selects whole
          posterior rows and SCRAPI2 uses one column per replicate."),
  answer("It is not wrong; a narrower interval is always better.",
         message = "A narrower interval that ignores real correlated
          uncertainty is falsely confident, not better."),
  allow_retry = TRUE
)
```

### Poor PIT sample sizes and biased GE

```{r pit-bias, exercise=TRUE, exercise.setup = "pit-bias-setup", exercise.lines=16}
# night_width(n) returns the 90% interval width of the nighttime-expanded adult
# total when each stratum has n PIT tags. Try n = 50 and n = 500 and watch the
# width shrink (more tags, tighter binomial rate).
night_width <- function(n, B = 400) {
  totals <- numeric(B)
  wc_by_strat <- tapply(sim_ad$wc$wc, sim_ad$wc$stratum, sum)
  for (b in seq_len(B)) {
    pn <- rbinom(3, n, sim_ad$truth$p_night) / n
    totals[b] <- sum((wc_by_strat / sim_ad$wc_prop) / (1 - pn))
  }
  diff(quantile(totals, c(.05, .95)))
}
c(n50 = night_width(50), n500 = night_width(______))
```

```{r pit-bias-setup}
# sim_ad is available
```

```{r pit-bias-hint}
# Just pass 500 as the second sample size: night_width(500).
```

```{r pit-bias-solution}
night_width <- function(n, B = 400) {
  totals <- numeric(B)
  wc_by_strat <- tapply(sim_ad$wc$wc, sim_ad$wc$stratum, sum)
  for (b in seq_len(B)) {
    pn <- rbinom(3, n, sim_ad$truth$p_night) / n
    totals[b] <- sum((wc_by_strat / sim_ad$wc_prop) / (1 - pn))
  }
  diff(quantile(totals, c(.05, .95)))
}
c(n50 = night_width(50), n500 = night_width(500))
```

```{r biased-ge, echo=TRUE}
# A GE posterior centered 15% too high: coverage of the true total falls,
# because a confidently wrong center cannot be fixed by a wider interval.
cover_ge <- function(bias, nsim = 80, B = 200) {
  hit <- logical(nsim)
  for (s in seq_len(nsim)) {
    d <- simulate_smolt_season(seed = s, B = B)
    tot <- vapply(seq_len(B), function(b) {
      pb <- d$weeks$rate * pmin(d$ge_draws[, b] * bias, 1)
      sum(rbinom(6, round(d$weeks$Tally / pb), pb) / pb)
    }, numeric(1))
    ci <- quantile(tot, c(.05, .95))
    hit[s] <- ci[1] <= d$truth$N_total && d$truth$N_total <= ci[2]
  }
  mean(hit)
}
c(calibrated = cover_ge(1.00), biased_high = cover_ge(1.15))
```

```{r q-diag}
question(
  "Across these diagnostics, which failures a wider interval can fix and which
   it cannot:",
  answer("A wider interval fixes small-sample noise (poor PIT sizes); it cannot
          fix a biased GE center or a cancelled resample weight, which move the
          interval to the wrong place.", correct = TRUE,
         message = "Correct. Centering errors (biased GE, SR-cancellation) are
          not a width problem; only genuine sampling noise is."),
  answer("A wider interval fixes all of them.",
         message = "It cannot rescue an interval centered on the wrong quantity;
          that is a bias, not a variance, problem."),
  allow_retry = TRUE
)
```

## Wrap-up

You rebuilt the Lower Granite escapement estimator from a plug-in expansion to
the full compound SCRAPI2 replicate, added GE and GSI posterior draws and the
rear-type split one at a time, verified `stock_A + stock_B == WildSmolts` in
every row, saw why uniform within-stratum resampling keeps the `1/SR` correction
honest where the original SCRAPI's SR-weighted resample cancelled it, walked the
adult side from SCOBI's `wc_prop = 1` to EASE's `5/6` with nighttime passage,
composition, and fallback separating ascensions from escapement, and diagnosed
the characteristic failure modes.

For the production map of what is resampled where, see `docs/bootstrap_map.md`.
For how to run the whole track, see `LEARNING_PATH.md`.
