---
title: "Estimator histories: SCOBI, SCRAPI, EASE, and SCRAPI2"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Rebuild the Lower Granite escapement bootstrap step by step against one
  known-truth simulated fishery, following the estimators from SCOBI and
  SCRAPI to EASE and SCRAPI2.
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# Source the shared helpers from the repository R/ directory. The tutorial runs
# from inst/tutorials/estimator-histories/, so R/ is three levels up; running
# from the repository root also works.
hp <- Filter(file.exists, c("../../../R/tutorial_helpers.R",
                            "R/tutorial_helpers.R",
                            "../../R/tutorial_helpers.R"))
if (length(hp) == 0) stop("tutorial_helpers.R not found")
source(hp[[1]])

# One simulated fishery, preserved across every module.
B_boot <- 1000L
alpha  <- 0.10
sim    <- simulate_smolt_season(seed = 6L, B = B_boot)
weeks  <- sim$weeks
fish   <- sim$fish
ge_draws  <- sim$ge_draws
gsi_draws <- sim$gsi_draws
truth  <- sim$truth
sim_ad <- simulate_adult_season(seed = 6L)
```

## The shared estimator

Both production tools at Lower Granite Dam, EASE for adults and SCRAPI2 for
smolts, run one estimator four times over. Written out, the skeleton is:

1. an observed **count** at the dam;
2. an **inclusion or detection probability** the count is only a fraction of;
3. an **expanded abundance**, count divided by that probability;
4. a **composition** split of the total into groups;
5. **uncertainty** propagated to an interval by resampling and posterior draws.

The whole tutorial builds each piece by hand against one simulated smolt season
whose truth we know. Here is the observed season.

```{r show-season, echo=TRUE}
weeks
truth$N_total   # true total wild smolts we are trying to recover
```

Each week has a raw trap `Tally`, a trap sample `rate`, and a guidance
efficiency `GE_true`. The inclusion probability for the count is
`Ptrue = rate * GE`, and abundance expands as `count / Ptrue`.

### Complete the expansion

Write the plug-in expansion: the total wild smolts implied by the observed
counts at the point (non-random) inclusion probability.

```{r expand, exercise=TRUE, exercise.lines=5}
# weeks has columns Tally, rate, GE_true. Inclusion prob is rate * GE_true.
# Fill in the expansion and compare it to the truth.
ptrue <- weeks$rate * weeks$GE_true
N_hat <- ______
c(N_hat = round(N_hat), truth = truth$N_total)
```

```{r expand-hint}
# Expanded abundance is the observed count divided by its inclusion
# probability, summed over weeks: sum(count / ptrue).
```

```{r expand-solution}
ptrue <- weeks$rate * weeks$GE_true
N_hat <- sum(weeks$Tally / ptrue)
c(N_hat = round(N_hat), truth = truth$N_total)
```

```{r q-expand}
question(
  "The plug-in estimate for this realized season lands somewhat above the true
   6800. What is the single most important reason a ratio estimator like
   count / (rate * GE) is unstable here?",
  answer("The low-GE stratum divides a handful of counts by a probability near
          0.01, so a few counted fish stand in for very many.", correct = TRUE,
         message = "Right: when GE is ~0.08 and the rate is ~0.12, Ptrue is
          ~0.01, and each counted fish expands to ~100. Small count noise
          becomes large abundance noise, which is exactly the right tail the
          later modules propagate."),
  answer("The trap sample rate is biased.",
         message = "The rate enters symmetrically as a known constant here; it
          is the tiny GE, not the rate, that drives the instability."),
  answer("The counts are simulated, so they cannot be trusted.",
         message = "The counts are an honest binomial draw from truth; the
          instability is real, not an artifact of simulation."),
  allow_retry = TRUE
)
```

**Locate this in production.** In `smoltEASE/R/SCRAPI2.R`, the inner estimator
`thetahat()` forms daily abundance as `passage$Tally / passage$Ptrue` and sums
by stratum. On the adult side, `escapeLGD/R/night_fall_reascend_wc_binom.R`
(`expand_wc_binom_night`) expands a window count `wc` by `1 / wc_prop` and by
`1 / (1 - p_night)`. Same skeleton, different names for the probability.

## Build a bootstrap from scratch

A point estimate is one number. The production interval comes from a *compound*
bootstrap: several independent sources of uncertainty stacked into one set of B
replicates. We add them one at a time and watch what each does to the interval.

### Stage 1: the plug-in, no uncertainty

```{r stage1, echo=TRUE}
ptrue     <- weeks$rate * weeks$GE_true
est_daily <- round(weeks$Tally / ptrue)   # implied true weekly passage
N_plugin  <- sum(weeks$Tally / ptrue)
round(N_plugin)
```

### Stage 2: the parametric passage-count bootstrap

The count itself is a binomial sample: of `est_daily` true passers, `Tally`
were counted, with probability `ptrue`. Resampling that binomial is the first
uncertainty source. Write one replicate.

```{r countboot, exercise=TRUE, exercise.setup = "countboot-setup", exercise.lines=12}
# est_daily and ptrue are defined for you. Draw B binomial resamples of the
# weekly counts and expand each back to a total. Report the bootstrap summary.
B <- 1000
totals <- numeric(B)
for (b in seq_len(B)) {
  cstar <- ______                       # rbinom over the 6 weeks
  totals[b] <- sum(cstar / ptrue)
}
round(boot_summary(totals, truth$N_total), 1)
```

```{r countboot-setup}
ptrue     <- weeks$rate * weeks$GE_true
est_daily <- round(weeks$Tally / ptrue)
```

```{r countboot-hint}
# rbinom is vectorised: rbinom(6, est_daily, ptrue) draws one resampled count
# per week in a single call. Put that inside the loop as cstar.
```

```{r countboot-solution}
B <- 1000
totals <- numeric(B)
for (b in seq_len(B)) {
  cstar <- rbinom(6, est_daily, ptrue)
  totals[b] <- sum(cstar / ptrue)
}
round(boot_summary(totals, truth$N_total), 1)
```

```{r q-countboot}
question(
  "With only the count bootstrap on, is the interval you get an honest 90%
   interval for total escapement?",
  answer("No: it captures count-sampling noise only, and ignores the
          uncertainty in GE, in composition, and in stock assignment, so it is
          too narrow.", correct = TRUE,
         message = "Correct. Each stage we add below widens the interval by a
          source the count bootstrap alone leaves out."),
  answer("Yes: resampling the count captures all the uncertainty there is.",
         message = "GE, composition, and GSI are all still fixed at point
          values here; their uncertainty has not entered yet."),
  allow_retry = TRUE
)
```

### Stage 3: one coherent GE posterior draw per replicate

GE is not known; it has a posterior. `ge_draws` is a 6-week by B matrix, each
**column** a coherent season-wide draw. The rule that matters: one replicate
uses one whole column, never a different column per week.

```{r geboot, exercise=TRUE, exercise.lines=14}
# ge_draws[, b] is this replicate's GE for all 6 weeks. weeks$rate and
# weeks$Tally are available. Build ptrue_b from column b, resample the count on
# that same basis, and expand.
B <- 1000
totals <- numeric(B)
for (b in seq_len(B)) {
  ptrue_b   <- weeks$rate * ge_draws[, b]
  est_b     <- round(weeks$Tally / ptrue_b)
  cstar     <- rbinom(6, est_b, ptrue_b)
  totals[b] <- ______
}
round(boot_summary(totals, truth$N_total), 1)
```

```{r geboot-hint}
# The expansion divides the resampled count by the SAME ptrue_b that generated
# it: sum(cstar / ptrue_b).
```

```{r geboot-solution}
B <- 1000
totals <- numeric(B)
for (b in seq_len(B)) {
  ptrue_b   <- weeks$rate * ge_draws[, b]
  est_b     <- round(weeks$Tally / ptrue_b)
  cstar     <- rbinom(6, est_b, ptrue_b)
  totals[b] <- sum(cstar / ptrue_b)
}
round(boot_summary(totals, truth$N_total), 1)
```

The distribution is now right-skewed: low-GE columns divide by a near-zero
probability and throw a long right tail.

```{r ge-skew, echo=TRUE}
set.seed(1)
totals <- vapply(seq_len(B_boot), function(b) {
  pb <- weeks$rate * ge_draws[, b]
  sum(rbinom(6, round(weeks$Tally / pb), pb) / pb)
}, numeric(1))
c(mean = mean(totals), median = median(totals), skew = skewness(totals))
```

```{r q-ge-skew}
question(
  "Why does adding GE uncertainty widen the interval mostly on the high side?",
  answer("Abundance is proportional to 1/GE, and 1/GE is convex, so low GE
          draws blow up more than high GE draws shrink.", correct = TRUE,
         message = "Yes. This convexity is the same reason E(1/GE) exceeds
          1/E(GE), which the diagnostics module returns to."),
  answer("The count bootstrap is skewed and GE just amplifies it.",
         message = "The binomial count bootstrap is nearly symmetric; the skew
          comes from the 1/GE transform, not the count."),
  allow_retry = TRUE
)
```

**Locate this in production.** `SCRAPI2()` precomputes `ge_day_mat`, a days-by-B
matrix, and inside the bootstrap loop uses `ge_day_mat[, b]` for replicate `b`
(`ptrue_b <- pass[, PASSrate] * ge_day_mat[, b]`). The columns are built by
`generate_ge_draws()` in `smoltEASE`, which selects **whole MCMC rows** so the
GE values on different days stay jointly consistent within a draw.

### Stage 4: composition and GSI, resampled per replicate

The genotyped `fish` sample splits the wild total into stocks A and B. Each fish
carries an inclusion weight `SR`; the estimator weights by `1/SR`. Composition
uncertainty comes from resampling the fish; GSI uncertainty comes from drawing a
fresh stock-assignment column per replicate (`gsi_draws`).

```{r additivity, exercise=TRUE, exercise.setup = "additivity-setup", exercise.lines=18}
# For one replicate: resample fish uniformly, draw GSI column b, estimate pA,
# and split a fixed wild total into stock A and stock B. Then check that the two
# stock abundances add back to the total WITHIN the replicate.
b <- 1
wild_total <- truth$N_total
idx   <- sample.int(nrow(fish), replace = TRUE)
stock <- gsi_draws[idx, b]
sr    <- fish$SR[idx]
pA_b  <- sum((1 / sr) * (stock == "A")) / sum(1 / sr)
stockA <- wild_total * pA_b
stockB <- wild_total * (1 - pA_b)
additivity_error <- ______              # should be ~0 by construction
c(stockA = round(stockA), stockB = round(stockB),
  sum = round(stockA + stockB), total = wild_total,
  additivity_error = additivity_error)
```

```{r additivity-setup}
b <- 1
```

```{r additivity-hint}
# pA_b + (1 - pA_b) = 1 exactly, so stockA + stockB - wild_total is 0 up to
# floating point: additivity_error <- (stockA + stockB) - wild_total.
```

```{r additivity-solution}
b <- 1
wild_total <- truth$N_total
idx   <- sample.int(nrow(fish), replace = TRUE)
stock <- gsi_draws[idx, b]
sr    <- fish$SR[idx]
pA_b  <- sum((1 / sr) * (stock == "A")) / sum(1 / sr)
stockA <- wild_total * pA_b
stockB <- wild_total * (1 - pA_b)
additivity_error <- (stockA + stockB) - wild_total
c(stockA = round(stockA), stockB = round(stockB),
  sum = round(stockA + stockB), total = wild_total,
  additivity_error = additivity_error)
```

```{r q-additivity}
question(
  "SCRAPI2 reports the bootstrap MEAN, not the median. Why does the mean keep
   stock A + stock B exactly equal to the wild total, while the median need
   not?",
  answer("Within every replicate the proportions sum to 1, so the stock
          abundances sum to the total row by row; averaging preserves a row-wise
          identity, taking medians column by column does not.", correct = TRUE,
         message = "Exactly. This is why SCRAPI2 fixes pointEst = 'mean': it is
          the only summary that keeps the reported components additive."),
  answer("The median is biased and the mean is unbiased.",
         message = "Additivity is about a linear identity holding row-wise, not
          about bias; the median of sums need not equal the sum of medians."),
  allow_retry = TRUE
)
```

## SCRAPI to SCRAPI2

The original SCRAPI and the current SCRAPI2 share this skeleton but differ in
three decisions: whether GE and GSI are fixed or uncertain, how fish are
resampled, and how the point estimate is defined. The most consequential is the
resampling choice, because it interacts with the `1/SR` weight already in the
estimator.

### Uniform vs SR-weighted resampling

The estimator weights each fish by `1/SR`. The question is whether the
*resample* should also use `SR`. Try both and measure coverage of the true
stock-A proportion across many simulated seasons.

```{r resample, exercise=TRUE, exercise.setup = "resample-setup", exercise.lines=22}
# coverage() re-simulates a season each iteration and builds a composition
# interval either uniformly or weighted by SR. Set `weighted` and compare.
coverage <- function(weighted, nsim = 150, B = 300) {
  hit <- logical(nsim)
  for (s in seq_len(nsim)) {
    d  <- simulate_smolt_season(seed = s, B = 1L)
    ff <- d$fish
    th <- numeric(B)
    for (b in seq_len(B)) {
      idx <- if (weighted)
        sample.int(nrow(ff), replace = TRUE, prob = ff$SR)
      else
        sample.int(nrow(ff), replace = TRUE)
      th[b] <- sum((1 / ff$SR[idx]) * (ff$stock[idx] == "A")) / sum(1 / ff$SR[idx])
    }
    ci <- quantile(th, c(.05, .95))
    hit[s] <- ci[1] <= d$truth$pA && d$truth$pA <= ci[2]
  }
  mean(hit)
}
c(uniform = coverage(FALSE), weighted = coverage(TRUE))
```

```{r resample-setup}
# nothing extra needed; simulate_smolt_season and stats are available
```

```{r resample-hint}
# You do not need to edit anything to see the effect: run it as is. The
# uniform arm keeps 1/SR honest; the weighted arm resamples proportional to SR
# AND divides by SR, so the two cancel.
```

```{r resample-solution}
coverage <- function(weighted, nsim = 150, B = 300) {
  hit <- logical(nsim)
  for (s in seq_len(nsim)) {
    d  <- simulate_smolt_season(seed = s, B = 1L)
    ff <- d$fish
    th <- numeric(B)
    for (b in seq_len(B)) {
      idx <- if (weighted)
        sample.int(nrow(ff), replace = TRUE, prob = ff$SR)
      else
        sample.int(nrow(ff), replace = TRUE)
      th[b] <- sum((1 / ff$SR[idx]) * (ff$stock[idx] == "A")) / sum(1 / ff$SR[idx])
    }
    ci <- quantile(th, c(.05, .95))
    hit[s] <- ci[1] <= d$truth$pA && d$truth$pA <= ci[2]
  }
  mean(hit)
}
c(uniform = coverage(FALSE), weighted = coverage(TRUE))
```

```{r q-resample}
question(
  "The uniform arm covers near 90%; the SR-weighted arm covers almost never.
   What has the SR-weighted resample actually estimated?",
  answer("The unweighted sample composition, because resampling proportional to
          SR cancels the 1/SR weight in the estimator, discarding the
          Horvitz-Thompson correction the weight exists to apply.",
         correct = TRUE,
         message = "Correct, and it is why stock A (sampled at a lower rate) is
          badly underestimated. The interval is centered on the wrong quantity,
          so it misses the truth almost every time."),
  answer("The same weighted composition, just with a wider interval.",
         message = "It is not merely wider; it is centered on a different
          quantity (the unweighted proportion), so its coverage collapses."),
  allow_retry = TRUE
)
```

**Locate this in production.** `smoltEASE/R/SCRAPI.R` (original) resamples fish
with `sample.int(nwk, replace = TRUE, prob = unlist(justwk$SR))` while its
`thetahat` still divides by `SR` -- the cancellation you just saw.
`smoltEASE/R/SCRAPI2.R` resamples with `sample.int(nrow(jw), replace = TRUE)`
(no `prob`) and keeps `1/SR` and `1/True` only inside the estimator. The
SCRAPI2 documentation states this directly: "weighting the resample by them
while also applying their inverse inside the estimator would tend to cancel the
intended inverse-probability correction."

### The point estimate is the bootstrap mean

SCRAPI2 does not report the raw data evaluated once. Every one of the B rows is
a full replicate, and the reported `Estimate` is the mean of those rows, while
`LCI`/`UCI` are percentiles of the same rows.

```{r mean-vs-raw, echo=TRUE}
set.seed(2)
totals <- vapply(seq_len(B_boot), function(b) {
  pb <- weeks$rate * ge_draws[, b]
  sum(rbinom(6, round(weeks$Tally / pb), pb) / pb)
}, numeric(1))
raw_point  <- sum(weeks$Tally / (weeks$rate * weeks$GE_true))
boot_mean  <- mean(totals)
pctl       <- quantile(totals, c(alpha/2, 1 - alpha/2), names = FALSE)
c(raw_point = round(raw_point), boot_mean = round(boot_mean),
  lci = round(pctl[1]), uci = round(pctl[2]))
```

```{r q-meanraw}
question(
  "For a strongly right-skewed cell, the bootstrap MEAN can fall OUTSIDE its own
   percentile interval. Is that a coding bug?",
  answer("No: mean and quantile are different summaries of the same skewed
          distribution, so the mean can sit above the 95th percentile.",
         correct = TRUE,
         message = "Right. SCRAPI2 flags such rows with a message rather than
          hiding them; it signals a heavy right tail to inspect, not an error."),
  answer("Yes: the point estimate must always be inside the interval.",
         message = "Only for symmetric summaries. A mean and a percentile of a
          skewed distribution are genuinely different functionals."),
  allow_retry = TRUE
)
```

## SCOBI to EASE

The adult estimator has its own history. Old SCOBI assumed the counting window
was open the whole time (`wc_prop = 1`); EASE corrects for the window being
closed part of each hour (`wc_prop = 5/6`), then expands for nighttime passage,
composition, and fallback. We use one simulated adult season.

```{r adult-season, echo=TRUE}
sim_ad$wc                 # weekly window counts
sim_ad$night              # nighttime-passage PIT data by stratum
sim_ad$fallback           # fallback/reascension PIT data by stratum
sim_ad$truth$N_total      # true adult escapement
```

### SCOBI vs EASE window-count assumption

```{r wcprop, exercise=TRUE, exercise.lines=8}
# The window is open a fraction wc_prop of the time, so the count must be
# expanded by 1/wc_prop. Compute the expanded daytime total under the old SCOBI
# assumption (wc_prop = 1) and under EASE (wc_prop = 5/6).
wc_total <- sum(sim_ad$wc$wc)
scobi <- wc_total / 1
ease  <- ______
c(raw = wc_total, scobi = round(scobi), ease = round(ease))
```

```{r wcprop-hint}
# EASE divides by 5/6, which multiplies the count by 6/5: ease <- wc_total / (5/6).
```

```{r wcprop-solution}
wc_total <- sum(sim_ad$wc$wc)
scobi <- wc_total / 1
ease  <- wc_total / (5/6)
c(raw = wc_total, scobi = round(scobi), ease = round(ease))
```

```{r q-wcprop}
question(
  "What does the old SCOBI wc_prop = 1 assumption imply, and why does EASE
   change it?",
  answer("SCOBI assumes every fish is counted (window always open); EASE
          recognizes the window is closed part of each hour, so the raw count is
          only 5/6 of daytime passage and must be expanded.", correct = TRUE,
         message = "Correct. wc_prop = 1 leaves out no daytime fish by
          assumption; 5/6 restores the ones passing while the window was shut."),
  answer("SCOBI expands more than EASE.",
         message = "The opposite: wc_prop = 1 applies no expansion at all, so
          SCOBI's daytime total is lower than EASE's."),
  allow_retry = TRUE
)
```

### Nighttime passage, composition, and fallback stacked

```{r adult-full, echo=TRUE}
set.seed(3)
B  <- B_boot
wc_prop <- sim_ad$wc_prop
wc_by_strat <- tapply(sim_ad$wc$wc, sim_ad$wc$stratum, sum)
nd <- sim_ad$night; fd <- sim_ad$fallback
totals <- numeric(B)
for (b in seq_len(B)) {
  # window count: parametric binomial bootstrap, expanded by 1/wc_prop
  wc_b <- vapply(seq_along(wc_by_strat), function(s)
    rbinom(1, round(wc_by_strat[s] / wc_prop), wc_prop) / wc_prop, numeric(1))
  # nighttime passage rate per stratum, binomial bootstrap, expand by 1/(1-p)
  pn_b <- rbinom(3, nd$totalPass, nd$nightPass / nd$totalPass) / nd$totalPass
  day_plus_night <- wc_b / (1 - pn_b)
  # net fallback rate per stratum, binomial bootstrap, remove reascension-net fallback
  pf_b <- rbinom(3, fd$totalPass, fd$numReascend / fd$totalPass) / fd$totalPass
  totals[b] <- sum(day_plus_night * (1 - pf_b))
}
round(boot_summary(totals, sim_ad$truth$N_total), 1)
```

**Locate this in production.** `escapeLGD/R/night_fall_reascend_wc_binom.R`
holds `expand_wc_binom_night()` (window-count binomial bootstrap `rbinom(boots,
round(wc/wc_prop), wc_prop)/wc_prop`, then `1/(1-p_night)`) and `nightFall()`
(binomial bootstrap of the nighttime rate, and a two-parameter fallback/
reascension likelihood via `optimllh` in `fallback_reascend_likelihood.R`).
Composition is `HNC_expand()` / `ascension_composition()`; the pieces are
multiplied together and turned into a percentile interval in
`apply_fallback_rates()` (`apply_fallback_rates_summarize.R`). The EASE annual
workflow (`EASE-code`, e.g. "SY2025 STHD Step 3") calls this with `wc_prop = 5/6`
and notes `wc_prop = 1` for old SCOBI.

```{r q-adult}
question(
  "Nighttime passage on the adult side is a bootstrapped binomial rate, not a
   fitted posterior like smolt GE. What is the practical consequence?",
  answer("It cannot be miscentered by a model: its uncertainty is only sampling
          noise in the PIT tags, so poor PIT sample sizes widen it but a wrong
          model cannot bias it the way a biased GE posterior biases SCRAPI2.",
         correct = TRUE,
         message = "Correct. This is the one genuine structural difference
          between the adult and smolt detection terms."),
  answer("It makes the adult interval always narrower than the smolt interval.",
         message = "Width depends on sample sizes and the other stacked
          sources; the point is about centering, not a guaranteed width order."),
  allow_retry = TRUE
)
```

## Model diagnostics and failure modes

The compound bootstrap has characteristic ways of going wrong. Each below is a
thing to check before quoting an interval.

### E(1/GE) is not 1/E(GE)

```{r jensen, echo=TRUE}
ge_col_means <- rowMeans(ge_draws)                 # per-week posterior mean GE
one_over_mean <- sum(weeks$Tally / (weeks$rate * ge_col_means))
mean_of_one_over <- mean(vapply(seq_len(B_boot), function(b)
  sum(weeks$Tally / (weeks$rate * ge_draws[, b])), numeric(1)))
c(expand_at_mean_GE = round(one_over_mean),
  mean_of_expansions = round(mean_of_one_over))
```

```{r q-jensen}
question(
  "The mean of the expansions exceeds the expansion at the mean GE. Which is the
   honest point estimate for total abundance?",
  answer("The mean of the expansions: abundance is a function of GE, so its
          expectation is the average of the function over the GE posterior, not
          the function evaluated at the average GE.", correct = TRUE,
         message = "Correct. SCRAPI2 computes exactly this by averaging the B
          replicate expansions, which is why it does not plug the posterior mean
          GE into one expansion."),
  answer("The expansion at the mean GE, because it uses one clean GE value.",
         message = "That is the biased one: by Jensen's inequality it
          understates the mean of a convex function of GE."),
  allow_retry = TRUE
)
```

### Mixing GE columns across weeks

The rule from module 2 was: one replicate, one whole GE column. Break it on
purpose and diagnose the damage.

```{r mix, exercise=TRUE, exercise.lines=16}
# WRONG on purpose: draw an INDEPENDENT GE column for each of the 6 weeks within
# a replicate, instead of one shared column. Compare the interval width to the
# coherent version.
B <- 1000
coherent <- mixed <- numeric(B)
for (b in seq_len(B)) {
  pc <- weeks$rate * ge_draws[, b]                       # coherent: one column
  coherent[b] <- sum(weeks$Tally / pc)
  cols <- ______                                         # 6 independent columns
  pm <- weeks$rate * ge_draws[cbind(1:6, cols)]
  mixed[b] <- sum(weeks$Tally / pm)
}
c(coherent_width = diff(quantile(coherent, c(.05,.95))),
  mixed_width    = diff(quantile(mixed,    c(.05,.95))))
```

```{r mix-hint}
# Pick 6 column indices independently: cols <- sample.int(ncol(ge_draws), 6,
# replace = TRUE). ge_draws[cbind(1:6, cols)] then takes week i from column
# cols[i].
```

```{r mix-solution}
B <- 1000
coherent <- mixed <- numeric(B)
for (b in seq_len(B)) {
  pc <- weeks$rate * ge_draws[, b]
  coherent[b] <- sum(weeks$Tally / pc)
  cols <- sample.int(ncol(ge_draws), 6, replace = TRUE)
  pm <- weeks$rate * ge_draws[cbind(1:6, cols)]
  mixed[b] <- sum(weeks$Tally / pm)
}
c(coherent_width = diff(quantile(coherent, c(.05,.95))),
  mixed_width    = diff(quantile(mixed,    c(.05,.95))))
```

```{r q-mix}
question(
  "Mixing independent GE columns across weeks makes the interval narrower. Why
   is that wrong?",
  answer("The weeks share strata and model parameters, so their GE values are
          correlated in the posterior; averaging independent draws cancels that
          shared uncertainty and understates the true spread.", correct = TRUE,
         message = "Correct. This is why generate_ge_draws() selects whole
          posterior rows and SCRAPI2 uses one column per replicate."),
  answer("It is not wrong; a narrower interval is always better.",
         message = "A narrower interval that ignores real correlated
          uncertainty is falsely confident, not better."),
  allow_retry = TRUE
)
```

### Poor PIT sample sizes and biased GE

```{r pit-bias, exercise=TRUE, exercise.setup = "pit-bias-setup", exercise.lines=18}
# night_width(n) returns the 90% interval width of the nighttime-expanded adult
# total when each stratum has n PIT tags. Try n = 50 and n = 500 and see the
# width shrink. Then set ge_bias = 1.15 to feed SCRAPI2 a GE posterior 15% too
# high and watch smolt coverage fall.
night_width <- function(n, B = 400) {
  totals <- numeric(B)
  wc_by_strat <- tapply(sim_ad$wc$wc, sim_ad$wc$stratum, sum)
  for (b in seq_len(B)) {
    pn <- rbinom(3, n, sim_ad$truth$p_night) / n
    totals[b] <- sum((wc_by_strat / sim_ad$wc_prop) / (1 - pn))
  }
  diff(quantile(totals, c(.05, .95)))
}
c(n50 = night_width(50), n500 = night_width(______))
```

```{r pit-bias-setup}
# helpers and sim_ad available
```

```{r pit-bias-hint}
# Just pass 500 as the second sample size: night_width(500). More tags means a
# tighter binomial rate and a narrower expanded interval.
```

```{r pit-bias-solution}
night_width <- function(n, B = 400) {
  totals <- numeric(B)
  wc_by_strat <- tapply(sim_ad$wc$wc, sim_ad$wc$stratum, sum)
  for (b in seq_len(B)) {
    pn <- rbinom(3, n, sim_ad$truth$p_night) / n
    totals[b] <- sum((wc_by_strat / sim_ad$wc_prop) / (1 - pn))
  }
  diff(quantile(totals, c(.05, .95)))
}
c(n50 = night_width(50), n500 = night_width(500))
```

```{r biased-ge, echo=TRUE}
# A GE posterior centered 15% too high: coverage of the true total falls,
# because a confidently wrong center cannot be fixed by a wider interval.
cover_ge <- function(bias, nsim = 80, B = 200) {
  hit <- logical(nsim)
  for (s in seq_len(nsim)) {
    d <- simulate_smolt_season(seed = s, B = B)
    tot <- vapply(seq_len(B), function(b) {
      pb <- d$weeks$rate * pmin(d$ge_draws[, b] * bias, 1)
      sum(rbinom(6, round(d$weeks$Tally / pb), pb) / pb)
    }, numeric(1))
    ci <- quantile(tot, c(.05, .95))
    hit[s] <- ci[1] <= d$truth$N_total && d$truth$N_total <= ci[2]
  }
  mean(hit)
}
c(calibrated = cover_ge(1.00), biased_high = cover_ge(1.15))
```

```{r q-diag}
question(
  "Across these diagnostics, which failures a wider interval can fix and which
   it cannot:",
  answer("A wider interval fixes small-sample noise (poor PIT sizes); it cannot
          fix a biased GE center or a cancelled resample weight, which move the
          interval to the wrong place.", correct = TRUE,
         message = "Correct. Centering errors (biased GE, SR-cancellation) are
          not a width problem; only genuine sampling noise is."),
  answer("A wider interval fixes all of them.",
         message = "It cannot rescue an interval centered on the wrong quantity;
          that is a bias, not a variance, problem."),
  allow_retry = TRUE
)
```

## Wrap-up

You rebuilt the Lower Granite escapement estimator from a plug-in expansion to
the full compound SCRAPI2 bootstrap, added GE and GSI posterior draws one at a
time, saw why uniform within-stratum resampling keeps the `1/SR` correction
honest where the original SCRAPI's SR-weighted resample cancelled it, walked the
adult side from SCOBI's `wc_prop = 1` to EASE's `5/6` with nighttime passage and
fallback, and diagnosed the characteristic failure modes.

For the production map of what is resampled where, see `docs/bootstrap_map.md`.
For how to run the whole track, see `LEARNING_PATH.md`.
